---
title: "Map"
author: "Alexander Eckerle"
date: today
format: html
editor: source
---

# Setup
```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(here)
library(conflicted)
library(janitor)
library(raster)
library(terra)
library(tidyterra)
library(rnaturalearth)
library(ggspatial)
library(sf)
library(ggnewscale)
library(formattable)
library(rgbif)
library(robis)

conflict_prefer_all("dplyr", quiet = TRUE)
```

# Loading TIFF map data
```{r}
#| label: load-tiffs
#| message: false
#| warning: false
#| echo: false

# Tiff files from https://kartkatalog.geonorge.no/metadata/kartverket/sjo-terrengmodeller-dtm-50/67a3a191-49cc-45bc-baf0-eaaf7c513549 and https://kartkatalog.geonorge.no/metadata/e25d0104-0858-4d06-bba8-d154514c11d2

bath1 <- rast(here("data/bc_50m/NHS-D2830_50M_E25833.tif"))
bath2 <- rast(here("data/bc_50m/NHS-D2831_50M_E25833.tif"))
bath3 <- rast(here("data/bc_50m/NHS-D2832_50M_E25833.tif"))
bath4 <- rast(here("data/bc_50m/NHS-D2833_50M_E25833.tif"))
bath5 <- rast(here("data/bc_50m/NHS-D2930_50M_E25833.tif"))
bath6 <- rast(here("data/bc_50m/NHS-D2931_50M_E25833.tif"))
bath7 <- rast(here("data/bc_50m/NHS-D2932_50M_E25833.tif"))
bath8 <- rast(here("data/bc_50m/NHS-D2933_50M_E25833.tif"))
topo <- rast(here("data/and_50m/7605_50m_33.tif"))
```

## Merging TIFF files
```{r}
#| label: merge-tiffs
#| message: false
#| warning: false
#| echo: false

bath_col <- sprc(bath1, bath2, bath3, bath4, bath5, bath6, bath7, bath8)

bath <- merge(bath_col)
```

### Changing chart date
```{r}
#| label: chart-wgs84
#| message: false
#| warning: false
#| echo: false

bath_wgs84 <- bath |>
   project("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
```

## Define map extent
```{r}
#| label: map-extent
#| message: false
#| warning: false
#| echo: false

bc_extent <- ext(15.5, 16.5, 69.3, 69.7)

bath_cropped <- crop(bath_wgs84, bc_extent)  
```

# Loading Zoe's sighting data
```{r}
#| label: load-sights-zoe
#| message: false
#| warning: false
#| echo: false

gps_zoe <- read_delim(here("data/Rissos_dolphins.csv"), 
                      delim = ",",
                      show_col_types = FALSE) |>
  clean_names() 
```

## Convert latitude and longitude - function
```{r}
#| label: func-conv-lat
#| message: false
#| warning: false
#| echo: false

lat_dec <- function(dms_string) {
  # Extract components
  deg <- parse_number(dms_string)
  min <- parse_number(gsub("^...", "", dms_string))
  
  # Convert to decimal degrees
  dec_deg <- deg + min / 60
  
  # Handle potential NA values
  result <- ifelse(is.na(dec_deg), NA, dec_deg)
  
  dec_deg
}
```

## Converting latitude and longitude
```{r}
#| label: conv-lat-long
#| message: false
#| warning: false
#| echo: false

gps_zoe$lat_dec <- lat_dec(gps_zoe$latitude_original)
gps_zoe$long_dec <- lat_dec(gps_zoe$longitude_original)  
#drop_na(gps$lat_dec, gps$long_dec)

# ATTENTION: this only works for longitudes shown with two digits for degrees 
# Format must be: "DD°M..."
```

## Adding year column
```{r}
#| label: add-year
#| message: false
#| warning: false
#| echo: false

gps_zoe$Year <- factor(year(gps_zoe$date))
```

# Loading my sighting data
```{r}
#| label: load-sights-alex
#| message: false
#| warning: false
#| echo: false

gps_alex <- read_delim(here("data/sightings_record.csv"),
                       show_col_types = FALSE) |>
  clean_names() |>
  filter(platform == "Whale2Sea" | platform == "Norwegian Orca Survey") 
```

## Making latitude and longitude readable 
```{r}
#| label: lat-lon-del
#| message: false
#| warning: false
#| echo: false

gps_alex$lat_dec <- as.numeric(
  substr(gps_alex$lat_n, 1, nchar(gps_alex$lat_n)-2))
gps_alex$long_dec <- as.numeric(
  substr(gps_alex$long_n, 1, nchar(gps_alex$long_n)-2))
gps_alex$Year <- factor(gps_alex$year)
```

# Combine Zoe's and my data
```{r}
#| label: cmb-sght-data
#| message: false
#| warning: false
#| echo: false

gps_zoe <- gps_zoe |>
  select(lat_dec, long_dec, Year) |>
  drop_na()

gps_alex <- gps_alex |>
  select(lat_dec, long_dec, Year) |>
  drop_na()

gps <- rbind(gps_zoe, gps_alex)
```

# Bleik Canyon 
## Basic map
```{r}
#| label: fig-basemap
#| fig-cap: "Map of the Bleik Canyon area."
#| message: false
#| warning: false
#| echo: false

map_base <- ggplot() +
  geom_spatraster(data = bath_cropped) +
  scale_fill_gradient(low = "dodgerblue4", 
                      high = "lightskyblue", 
                      na.value = "beige") +
  geom_rect(aes(xmin = 15.93,
                xmax = 15.99,
                ymin = 69.3,
                ymax = 69.325),
                colour = "lightskyblue", fill = "lightskyblue") +
  geom_spatraster_contour(data = bath_cropped,
                          colour = "grey70") +
  geom_text(aes(x = 16.05, y = 69.408, label = "200"), 
            colour = "grey60",
            size = 2.5,
            angle = 350) + 
  geom_text(aes(x = 15.798, y = 69.5, label = "400"),
            colour = "grey61",
            size = 2.5,
            angle = 80) +
  geom_text(aes(x = 15.793, y = 69.52, label = "600"),
            colour = "grey62",
            size = 2.5,
            angle = 60) +
  geom_text(aes(x = 15.8, y = 69.54, label = "800"),
            colour = "grey63",
            size = 2.5,
            angle = 70) +
  geom_text(aes(x = 15.802, y = 69.56, label = "1000"),
            colour = "grey64",
            size = 2.5,
            angle = 70) +
  geom_text(aes(x = 15.797, y = 69.58, label = "1200"),
            colour = "grey65",
            size = 2.5,
            angle = 75) +
  geom_text(aes(x = 15.727, y = 69.55, label = "1400"),
            colour = "grey66",
            size = 2.5,
            angle = 65) +
  geom_text(aes(x = 15.719, y = 69.57, label = "1600"),
            colour = "grey67",
            size = 2.5,
            angle = 55) +
  geom_text(aes(x = 15.678, y = 69.59, label = "1800"),
            colour = "grey68",
            size = 2.5,
            angle = 75) +
  geom_text(aes(x = 15.533, y = 69.593, label = "2000"),
            colour = "grey69",
            size = 2.5,
            angle = 15) +
  geom_text(aes(x = 16.33, y = 69.445, label = "400"),
            colour = "grey61",
            size = 2.5,
            angle = 20) +
  theme_bw() +
  scale_x_continuous(expand = c(0, 0),
                     breaks = c(15.6, 15.8, 16, 16.2, 16.4)) +
  scale_y_continuous(expand = c(0, 0),
                     breaks = c(69.4, 69.5, 69.6)) +
  annotation_scale(location = "br") +
  annotation_north_arrow(height = unit(0.8, "cm"),
                         width = unit(0.8, "cm"),
                         location = "tr") +
  geom_text(aes(x = 15.7, y = 69.45, label = "Bleik\nCanyon")) +
  geom_point(aes(x = 16.1194, y = 69.3144)) +
  geom_text(aes(x = 16.1194, y = 69.328, label = "Andenes")) +
  theme(#legend.position = "bottom",
        #legend.direction = "horizontal",
        axis.title = element_blank()) + #,
#        axis.text.y = element_text(angle = 90, hjust = 0.5)) +
  labs(fill = "Depth (m)") +
  #guides(fill = guide_colourbar(title.position = "top")) +
  NULL

map_base
```

# North East Atlantic
## Loading IUCN distribution
```{r}
#| label: load-iucn
#| message: false
#| warning: false
#| echo: false

iucn <- read_sf(here("data/IUCN/data_0.shp"))
```

## Landmass 
```{r}
#| label: land
#| message: false
#| warning: false
#| echo: false

europe <- ne_countries(scale = 50, continent = "Europe") 
```

## NEA map
```{r}
#| label: fig-nea-distr
#| fig-cap: "Distribution range vs. study area."
#| message: false
#| warning: false
#| echo: false

distr_map <- ggplot() +
  geom_sf(data = europe) +
  geom_sf(data = iucn, fill = "dodgerblue") +
  geom_rect(aes(xmin = bc_extent[1],
            xmax = bc_extent[2],
            ymin = bc_extent[3],
            ymax = bc_extent[4]),
            colour = "orange", fill = "orange", alpha = 0.5) + 
  coord_sf(xlim = c(-10, 25), ylim = c(55, 71)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank(), 
        plot.background = element_blank()) +
  NULL

distr_map
```
# BC + NEA combined map
```{r}
#| label: fig-cmb-maps
#| fig-cap: "Inlet map."
#| message: false
#| warning: false
#| echo: false

map_comb <- map_base + 
  annotation_custom(ggplotGrob(distr_map), 
                    xmin = 16.0, xmax = 16.49, ymin = 69.425, ymax = 69.6)

map_comb
```

# W2S sightings combined map 
```{r}
#| label: fig-sght
#| fig-cap: "Risso's dolphins sightings at the Bleik Canyon."
#| message: false
#| warning: false
#| echo: false

map_sight <- map_comb +
  geom_point(aes(x = long_dec, y = lat_dec, colour = Year), 
             data = gps, size = 0.5) +
  new_scale_fill() +
  scale_colour_brewer(palette = "YlOrRd") +
  theme(axis.title = element_blank()) +
  NULL

# legend is slightly edited using Inkscape to include a legend for inset map

map_sight
```